Follow-up Samples

### 2016-12-16 submission/manuscript_outline_20161024.Rmd

Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is the second most common cancer death among men aged 40-79 years of age [@jemal_cancer_2010; @haggar_colorectal_2009].  Over the last few years death due to the disease has seen a significant decrease thanks mainly to improvements in screening [@jemal_cancer_2010].  However, despite this giant improvement there are still approximately 50,000 deaths from the disease a year [@haggar_colorectal_2009].  It is estimated that around 5-10% of all CRCs can be explained by autosomal dominant inheritance [@green_very_2007]. The vast majority of CRCs are not inherited and the exact etiology to disease has not been well worked out [@haggar_colorectal_2009].

Recent reports have found that the microbiome between those with and without CRC are different and that bacteia associated with CRC tissue is different than that found on the normal mucosa (**Insert citations**).  Further, there is evidence to suggest that more inflammatory mouth-associated bacteria are more prevalent in the guts of those with CRC (**insert citation**).  This has led to the hypothesis that these bacteria replace traditional residents that produce metabolites such as butyrate (**insert citation**) and lead to a more inflammatory state (**insert citation**).  This inflammatory state allows for more of these types of bacteria to populate the gut and leads to neoplasia and eventually CRC (**insert citation**).

Building upon these findings a number of groups, including our own, have investigated using the bacterial microbiome as a potential biomarker for adenoma and CRC detection (**insert citations**).  These collectively show that a reasonable area under the curve (AUC) can be obtained when using the stool bacterial microbiome to classify disease state (**insert citation**).  Further, a number of groups have extended this observation to show that using either specific bacterial species or the bacterial microbiome as a whole paired with the Fecal Immunochemical Test (FIT) can increase the AUC of the model (**insert citations**).

In this study we further refine the random forest model by including more background information on the individuals within it. We used a rigourous and accepted approach for machine learning training and testing to validate this model (**insert citation**).  We then applied it to individuals in which we have samples before and after treatment to assess how well the model adjusts their probability of having an adenoma or carcinoma.  Finally, we investigated which specific Operational Taxonomic Units (OTUs) are most affected by treatment.  Within this study lesion refers to both adenoma and carcinoma.

>>>>

Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is the second most common cancer death among men aged 40-79 years of age [@jemal_cancer_2010; @haggar_colorectal_2009].  Over the last few years death due to the disease has seen a significant decrease thanks mainly to improvements in screening [@jemal_cancer_2010].  However, despite this giant improvement there are still approximately 50,000 deaths from the disease a year [@haggar_colorectal_2009].  It is estimated that around 5-10% of all CRCs can be explained by autosomal dominant inheritance [@green_very_2007]. The vast majority of CRCs are not inherited and the exact etiology to disease has not been well worked out [@haggar_colorectal_2009].  Although many risk factors have been identified [@hagger_colorectal_2009] and non-invasive screening techniques have started to be put into consistent use [@liao_application_2013; @johnson_multi-target_2016] there has been a consistent increase in the incidence of CRC in the younger population.

This increased incidence of CRC in the younger population is concerncing since having either an adenoma or carcinoma increases ones risk for future adenomas or carcinomas [@laiyemo_short-_2013; @matsuda_five-year_2009; @ren_long-term_2016].  This increased risk can also carry with it an increased risk of mortality due to this recurrence [@loberg_long-term_2014; @freeman_natural_2013].  Therefore there has been a great amount of interest in early risk stratification tools [@lee_identification_2016; @richards_evidence-based_2016] that can help identify those they may be at risk of reccurence.  Concurrently with this there has been a lot of interest in new areas such as the gut bacterial microbiome for insight into potential disease pathology.

There has been promising work on the bacterial microbiome and it's ability to be able to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014]. There has also been research into how this microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  There have also been a few studies that have shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved ith the pathogenesis of CRC.  These studies have helped to provide a tantilzing link between the bacterial microbiome and CRC. However, at this present time there remains limited information on the bacterial microbiome before and after successful surgery for removal of the adenoma or carcinoma.

In this study we investigated what happened to the bacterial microbiome before and after surgery for both adenoma and carcinoma individuals.  Our anlaysis includes both alpha and beta diversity analysis along with investigating indvidual operational taxonomic units (OTUs).  We next utilized a Random Forest model that was trained on a completely separate data set and observed how this model as well as specific OTUs within this model performed on the initial and follow up samples.  We then created a new model to classify intial and follow up samples and investigated how the model performs and what specific OUTs within the model change.  Finally, we observed how specific bacteria that have been previously implicated in CRC changed before and after treatment in both the adenoma and carcinoma individuals.
